\chapter{Software Testing and Prefix Notation}
\label{ch:software-testing-prefix-notation}

%\todo{ COMMENT ONLY, NO TODO Rex16Sep2017
%Removed reciprocals stuff
%see on Rex's local copy if needed: ch03WITHRECIPROCALSSTUFFINCASEWEREVERTTOIT16Sep2017.tex
%}

We will be using a software development environment that automates parts of the software testing process.
To use it, we can define tests that check results corresponding to inputs provided directly in the test.
Or, we can ask the system to generate random data and
check results against logic formulas specifying properties that we expect the results to have.

\begin{aside}{aside:proof-pad}{Proof Pad and ACL2}
Proof Pad is a tool that provides a convenient way
to write, test, and reason about programs
written in the ACL2 dialect of a programming language called Lisp.
ACL2 includes a mechanized logic that
supports rigorous checking of the reasoning process,
and we will use it for that purpose starting in this chapter.

We will begin to use \index{Proof Pad}Proof Pad in this chapter,
so you should install it on your computer.
Get it at the following URL. It's free.\\
\hspace*{38mm}\url{http://proofpad.org}
%\caption{Proof Pad and ACL2}
%\label{aside:proof-pad}
\end{aside}

To get the syntax straight,
let's specify some direct tests that Proof Pad
can carry out for the numeric addition operator.
The following tests check some equations that we expect to be true:
$2 + 2 = 4$, $5 + 7 = 10 + 2$, etc.

\index{teachpacks!:dir}\index{directory (:dir)!:teachpacks}\index{book!directory (:dir)}\index{teachpacks!testing}\index{testing!teachpack}\index{book!testing}\index{check-expect test}\index{testing!check-expect}
\label{testing-include}
\begin{code}
\begin{verbatim}
(include-book "testing" :dir :teachpacks)
(check-expect (+ 2 2) 4)
(check-expect (+ 5 7) (+ 10 2))
(check-expect (+ 27 6) (+ (+ 23 5) (+ 2 2)))
\end{verbatim}
\end{code}

This notation looks strange, with formulas like \textsf{(+ 2 2)} instead of $2 + 2$.
ACL2 expresses formulas with a
\index{prefix notation}\index{operator!prefix notation}prefix notation
rather than the customary \index{infix notation}\index{operator!infix notation}infix notation.
In ACL2's prefix notation, a computational
\index{formula!prefix notation}formula starts with a parenthesis,
followed by the operator it invokes.
The operator comes first (as a ``prefix''), then the operands follow.
Infix notation places the operator between the operands
but prefix notation puts it before the operands.

That seems simple enough, but it gets gradually more unfamiliar
as the formulas get more complicated.
For example, a formula for the sum of $x$ and three times $y$, $(x + 3*y)$,
comes out as \textsf{(+ $x$ ($*$ 3 $y$))} in prefix form.
This will take a little getting used to,
but it is a lot easier to get good at than
most of the stuff you have been studying.
We hope you will be willing to accept it as just another way of writing formulas.
One advantage is that the formulas are, by default, fully parenthesized,
so there is never any question about what operands go with which operators.
Some people end up liking the prefix format better than infix.
We will use both, in different contexts, depending on whether
we're doing paper-and-pencil reasoning or formal, mechanized logic.

Another important element is the
\seeonlyindex{include-book}{directive}\index{directive!include-book}\textsf{include-book}
directive.
The facility for tests of this kind resides in an ACL2 book
known as \textsf{testing}, which in turn resides in a directory known
\index{teachpacks!:dir}\index{directory (:dir)!:teachpacks}\index{book!directory (:dir)}\index{teachpacks!testing}\index{testing!teachpack}\index{book!testing}as
\textsf{:teachpacks} (with a leading colon).
The \textsf{include-book} directive tells the system
the name of the testing package and where to find it.
This boilerplate is needed (including the keyword \textsf{:dir}, with the leading colon)
whenever you want to use \textsf{check-expect} testing,
even though we won't repeat it in all of our examples. %'

If you install Proof Pad and put these tests in a file
with a .lisp extension (plusoptests.lisp, for example),
you can use Proof Pad to run the tests.
The first tests two pass without fanfare,
but the third test has a mistake.
We should not expect that $27 + 6 = (23 + 5) + (2 + 2)$.
The system reports that the assertion failed.
If you change the 6 to 5, the test will pass, as it should.
Normally, when we run tests, we want to know whether the
operator is doing the right thing,
but in this case the operator (\textsf{+}) was fine.
It was the test that was wrong.

Direct, one-off, \textsf{check-expect} tests
are a good way to get started, but
we will want to test more general properties that we specify with logic formulas.
Proof Pad facilitates that kind of \index{testing!doublecheck}testing
with a package called
\index{DoubleCheck}\index{book!doublecheck}\index{teachpacks!doublecheck}DoubleCheck
that generates
\index{random data}\index{testing!with random data}random data
to run against a predicate
(that is, a formula representing a true/false value).
To demonstrate the idea, consider testing the addition operator to see if
it conforms to the associative law for addition: $(x + (y + z)) = ((x + y) + z)$
(\{+ associative\}, figure~\ref{fig-02-01}, page \pageref{fig-02-01}).
To do that, we define a property as an equation
and direct Proof Pad to generate some random numbers and
report whether or not the equation holds for all of the random data.
The Proof Pad ``doublecheck'' package,
like the ``testing'' package,
resides in the
\index{teachpacks!:dir}\index{teachpacks!doublecheck}teachpacks directory.
An \textsf{include-book} directive is required to make it accessible for use.

\index{DoubleCheck}\index{testing!doublecheck}\index{book!doublecheck}\index{teachpacks!doublecheck}\index{teachpacks!:dir}\index{directory (:dir)!:teachpacks}\index{book!directory (:dir)}\index{testing!teachpack}\index{book!testing}\index{operator!numeric order ($<$, $<=$, $=$, $>=$, $>$)}\index{order!numeric ($<$, $<=$, $=$, $>=$, $>$)}\index{predicate!numeric order ($<$, $<=$, $=$, $>=$, $>$)}\index{number!order ($<$, $<=$, $=$, $>=$, $>$)}\index{compare!numbers ($<$, $<=$, $=$, $>=$, $>$)}\index{definition!property}\index{data, random test}\index{random data}
\label{dblchk-include}
\begin{code}
\begin{verbatim}
(include-book "doublecheck" :dir :teachpacks)
(defproperty +associative-test
  (x :value (random-integer)
   y :value (random-integer)
   z :value (random-integer))
  (= (+ x (+ y z))
     (+ (+ x y) z)))
\end{verbatim}
\end{code}

A \index{property!doublecheck}property definition with
DoubleCheck begins with a parenthesis, then the keyword
\index{testing!with random data}\index{random data}\index{defproperty}\index{property!defproperty}\index{data, random test}\textsf{defproperty}
and a name for the property to be tested,
which is \textsf{+associative-test} in this example.\footnote{The
rules for naming things in ACL2 are lax compared to many
programming languages. Names can be made up of letters and digits, of course,
but they can also include some special characters,
such as plus signs, hyphens, and colons (not semicolons or commas, though).}
A sequence of data generator specifications follows the name of the property.
The entire sequence is enclosed in parentheses,
and each specification has three parts: a name,
the keyword \textsf{:value} (with the leading colon),
and a generator, also enclosed in parentheses.
In this example, the definition says to use random integers.
DoubleCheck is able to generate many kinds of random data,
and we will gradually introduce different data generators as the story progresses.
Finally, a predicate formula specifies the test.
In this case, the predicate is an equation,
written of course in prefix notation.

Let's look at another example.
ACL2 can deal with a collection of data elements
by keeping them in an ordered \index{sequence}sequence
called, and this is a technical term, a
\index{list}\emph{list}.
There is an operator,
\label{list-op-informal}\index{operator, by name!list}\index{list!operator (\emph{see also} operator)}\textsf{list},
that creates a list of its operands
(any number of operands).
Another operator,
\label{append-op-informal}\index{operator, by name!append}\index{append!operator}\textsf{append},
\index{concatenate}concatenates two lists.
A few \textsf{check-expect} tests may clarify
what the \textsf{append} and \textsf{list} operators do.
Later, we will describe them formally.

\begin{code}
\begin{verbatim}
(check-expect (append (list 1 2 3 ) (list 4 5)) (list 1 2 3 4 5))
(check-expect (append (list 9 8 7) (list 6) (list 9 8 7 6))
(check-expect (append (list 11 7) (list 2 5 3)) (list 11 7 2 5 3))
(check-expect (append (list 2 0) (list 1 8)) (list 2 0 1 8))
\end{verbatim}	
\end{code}

We would anticipate that the length of a concatenation of two lists
would be the length of the first list plus the length of the second.
Using the
\label{len-op-informal}\index{operator, by name!len (length of list)}\seeonlyindex{len}{operator}\textsf{len}
operator, which computes the length of a list,
we can state this property in a DoubleCheck test.
Later, we will prove this relationship between \textsf{append} and \textsf{len},
but for the moment we confine ourselves to stating properties
and asking Proof Pad to check them against random data.

\label{additive-lengths-test}\index{definition!property}\index{data, random test}\index{random data}\index{defproperty}
\begin{code}
\begin{verbatim}
(defproperty additive-law-of-concatenation-test
  (xs :value (random-list-of (random-integer))
   ys :value (random-list-of (random-integer)))
  (= (len (append xs ys))
     (+ (len xs) (len ys))))
\end{verbatim}
\end{code}

\begin{aside}{natural-number-def}{Natural Numbers}
A \index{natural number}\emph{natural number} is
an integer that is zero or more.
Some treatments start the naturals with one
rather than zero, but in the computing domain,
starting with zero simplifies some formulas.
%\caption{Natural Numbers}
%\label{natural-number-def}
\end{aside}

When we discuss computer arithmetic, we will need to know something about
arithmetic on finite sets of numbers,  and the
\index{operator, by name!mod (remainder)}\label{mod-function}\textsf{mod}
operator will play a crucial role.
The \textsf{mod} operator takes two operands and
delivers the remainder in the division of the first operand by the second.
ACL2 does not restrict the operands to the set of natural numbers,
but in this discussion, we are going to stick with that domain.
Think
\index{divisor}divisor, \index{dividend}dividend, \index{quotient}quotient, \index{remainder}remainder,
\index{division!third grade}\index{division!long division}\index{division!floor, mod}as in
\seeonlyindex{third-grade long division}{division}long division:
third-grade stuff.

Let's test drive the \textsf{mod} operator, starting with
some simple checks, such as using \textsf{mod}
to compute the remainder when dividing by two,
which will produce zero for even numbers and one for odd numbers.

\begin{code}
\begin{verbatim}
(check-expect (mod 12 2) 0)
(check-expect (mod 27 2) 1)
\end{verbatim}
\end{code}

Here are a few more sanity checks, this time with three as the divisor:

\begin{code}
\begin{verbatim}
(check-expect (mod 14 3) 2)
(check-expect (mod  7 3) 1)
(check-expect (mod 18 3) 0)
\end{verbatim}
\end{code}

\begin{aside}{modular-arithmetic}{Modular Arithmetic}
\index{arithmetic!clock}\index{clock arithmetic}\index{arithmetic!modular}\index{modular arithmetic}\index{operator, by name!mod (remainder)}\index{division!third grade}\index{division!long division}\index{modular arithmetic}\index{operator, by name!mod (remainder)}\index{arithmetic!clock}\index{clock arithmetic}\index{arithmetic!modular}\index{division!mod (remainder)}Modular
arithmetic (aka clock arithmetic)
produces integers in a fixed range: $0$, $1$, $\dots$ $(m - 1)$,
where $m > 0$ is a natural number known as the
\index{modulus (modular arithmetic)}\emph{modulus}.
If $x$ is an integer, the formula ($x$ mod $m$) denotes
the remainder in the division of $x$ by $m$
($0 \leq (x$ mod $m) < m$).
Modular addition, subtraction, and multiplication
are consistent with ordinary arithmetic.\\
\hspace*{15mm}($(x + y)$ mod $m$) = ((($x$ mod $m$) $+$ ($y$ mod $m$)) mod $m$) \\
\hspace*{15mm}($(x - y)$ mod $m$) = ((($x$ mod $m$) $-$ ($y$ mod $m$)) mod $m$) \\
\hspace*{15mm}($(x \times y)$ mod $m$) = ((($x$ mod $m$) $\times$ ($y$ mod $m$)) mod $m$)\\
The ACL2 operation
(\textsf{mod} $x$ $m$) delivers the remainder in $x \div m$.
Our discussion restricts the operands to natural numbers
but ACL2 doesn't.
%\caption{Modular Arithmetic}
%\label{modular-arithmetic}
\end{aside}

We can use Proof Pad to put \textsf{mod}
through its paces on a large number of tests using DoubleCheck,
but for that we will need to come up with relationships
that express more general properties of division and remainders.
One such property is that the remainder doesn't change
when the divisor is added to the dividend.

\index{definition!property}\index{data, random test}\index{random data}\index{defproperty}
\begin{code}
\begin{verbatim}
(defproperty mod-invariant
  (divisor-minus-1 :value (random-natural)
   dividend        :value (random-natural))
  (let* ((divisor (+ divisor-minus-1 1))) ; avoid zero divisor
    (= (mod dividend divisor)
       (mod (+ dividend divisor) divisor))))
\end{verbatim}
\end{code}

Generating
\index{random data}\index{testing!with random data}random data
is an art.
In this example, we have made sure the divisor isn't zero
by adding one to a natural number.
Since negative numbers aren't natural numbers, adding one
to a natural number ensures that the sum is nonzero.
\label{let-example-brief-explain}
The definition of the \textsf{mod-invariant} property
uses a \textsf{let$*$} formula,
which provides a way to attach names to values temporarily
(box \ref{let-star-def}, page \pageref{let-star-def}).
In this case, the \textsf{let$*$} says that
the name \textsf{divisor} will stand for
the value represented by the formula \textsf{(+ divisor-minus-1 1)}.

\begin{aside}{let-star-def}{Local Names for Values: \textsf{let$*$}}
A
\index{formula!let$*$}\index{variable!let$*$}\index{let$*$, local name}\index{value, let$*$ name for}\index{local names (let$*$)}\index{value, let$*$ name for}\index{name, local (let$*$)}\textsf{let$*$}
formula attaches values to names.
The \index{scope!let$*$ name}scope of the \textsf{let$*$} opens with a parenthesis,
then the keyword \textsf{let$*$}.
The scope continues with
a sequence of name/value bindings
and ends with a formula for the value
to be delivered by the \textsf{let$*$}.
A final parenthesis closes the scope.

Each binding has the form ($h$ $v$),
where $h$ is a name and $v$ is a
formula specifying a value to be associated with that name
($h$ does not denote an operator).
The parentheses around each binding are required.
Any of the names attached to values in the bindings
denote those values at any
subsequent point within the scope
of the \textsf{let$*$},
but the name/value associations do not
extend outside that scope.
There can be any number of bindings.
The most common use of \textsf{let$*$} is to give names
to values needed more than once in a computation,
but sometimes \textsf{let$*$} is used to provide mnemonic names
to make a formula more easily understood.

The parentheses in a \textsf{let$*$} formula are tricky.
Each individual binding is enclosed in parentheses,
and the whole sequence of bindings is enclosed in another set
of parentheses. In addition, the entire \textsf{let$*$} formula
is enclosed in parentheses.
A \textsf{let$*$} formula delivers a value $v_{let*}$
specified by a formula positioned after name/value bindings.
%(and of course before the parenthesis that closes the \textsf{let$*$}).
\begin{center}
(let$*$ (($h_1$ $v_1$) ($h_2$ $v_1$) $\dots$ ($h_n$ $v_n$)) $v_{let*}$)
\end{center}
%\caption{Local Names for Values: \textsf{let$*$}}
%\label{let-star-def}
\end{aside}

Since (\textsf{mod} $x$ $m$) $<$ $m$ when $m > 0$,
the \textsf{mod} operator will pass the following test:

\index{compare!numbers ($<$, $<=$, $=$, $>=$, $>$)}\index{predicate!numeric order ($<$, $<=$, $=$, $>=$, $>$)}\index{order!numeric ($<$, $<=$, $=$, $>=$, $>$)}\index{operator!numeric order ($<$, $<=$, $=$, $>=$, $>$)}\index{number!order ($<$, $<=$, $=$, $>=$, $>$)}\seeonlyindex{less than ($<$)}{predicate}\index{definition!property}\index{data, random test}\index{random data}\index{defproperty}
\begin{code}
\begin{verbatim}
(defproperty mod-upper-limit-test
  (divisor-minus-1 :value (random-natural)
   dividend        :value (random-natural))
  (let* ((divisor (+ divisor-minus-1 1))) ; avoid zero divisor
    (< (mod dividend divisor) divisor)))
\end{verbatim}
\end{code}

In this test, the property is not expressed as an equation
but as an inequality
specified with the less-than operator ($<$).
As always, the formula puts the operator in the prefix position,
in front of its operands.
For practice, add this property to the .lisp file with the other tests and
use Proof Pad to run it.

When one natural number is divided by another,
the remainder is also a natural number.
We can use the logical-and operator
to combine the upper-limit test with the natural-number test
in one property definition
(``\textsf{and}'' is the ACL2 name for $\wedge$).
\label{natp-op}The value of the formula
\textsf{(natp $x$)} is true
if $x$ is a natural number and false if it isn't.

\label{natp-axiom-formal}
\begin{center}
\begin{tabular}{c}
Axiom \{\emph{natp}\} \\
\hline
(natp $x$) $=$ $x$ \emph{is a natural number} $=$ \textsf{(and (integerp $x$) (>= $x$ 0))}\\
\end{tabular}\\
\emph{Note}: \textsf{(integerp $x$)} is true if $x$ a whole number and false otherwise.
\end{center}

\index{definition!property}\index{data, random test}\index{random data}\index{defproperty}
\begin{code}
\begin{verbatim}
(defproperty mod-range-test
  (divisor-minus-1 :value (random-natural)
   dividend        :value (random-natural))
  (let* ((divisor (+ divisor-minus-1 1))) ; avoid zero divisor
    (and (natp (mod dividend divisor))
         (< (mod dividend divisor) divisor))))
\end{verbatim}
\end{code}

There are two parts to the result
of dividing one number by another: the quotient and the remainder.
The \textsf{mod} operator delivers the remainder
and an operator called \textsf{floor} (box~\ref{floor-ceiling-ops-brackets})
delivers the quotient.
The quotient is always strictly smaller
than the dividend when the divisor is bigger than one
and the dividend is a nonzero, natural number.
The following test checks for that property.
The random-value generator for the divisor
makes sure the divisor exceeds one by adding two
to a natural number.
Similarly, we make sure the dividend isn't zero by adding one.

\label{quotient-less-than-dividend-test}\index{definition!property}\index{data, random test}\index{random data}\index{defproperty}
\begin{code}
\begin{verbatim}
(defproperty quotient-less-than-dividend-test
  (divisor-minus-2   :value (random-natural)
   dividend-minus-1  :value (random-natural))
  (let* ((divisor  (+ divisor-minus-2 2))   ; divisor  > 1
         (dividend (+ dividend-minus-1 1))) ; dividend > 0
    (< (floor dividend divisor)
       dividend)))
\end{verbatim}
\end{code}

Checking the result of a division is a matter
of multiplying the quotient by the divisor and adding the remainder.
If this fails to reproduce the dividend,
something has gone wrong in the division process.
The following property tests this relationship
between the \textsf{mod} and \textsf{floor} operators.
It needs to use the multiplication operator ($*$).

\label{division-check-test}
\begin{code}
\begin{verbatim}
(defproperty division-check-test
  (divisor-minus-1 :value (random-natural)
   dividend        :value (random-natural))
  (let* ((divisor (+ divisor-minus-1 1))) ; avoid zero divisor
    (= (+ (* divisor (floor dividend divisor))
          (mod dividend divisor))
       dividend)))
\end{verbatim}
\end{code}

We hope by now you are starting to get comfortable with prefix notation
and using Proof Pad to run tests.
The exercises will give you a chance to practice.

\begin{exercises}
	
\exer {Define a test of the \textsf{floor} operator
that checks to make sure its value is a natural number
when its operands are natural numbers
and the divisor (second operand) is not zero.
Use Proof Pad to run the test.}

\exer {The
\index{operator, by name!max (of two operands)}\seeonlyindex{max}{operator}\textsf{max}
operator chooses the larger of two numbers:
\textsf{(max 2 7)} is \textsf{7}, \textsf{(max 9 3)} is \textsf{9}.
Define a DoubleCheck property that tests to make sure
\textsf{(max $x$ $y$)} is
\index{compare!numbers ($<$, $<=$, $=$, $>=$, $>$)}\index{predicate!numeric order ($<$, $<=$, $=$, $>=$, $>$)}\index{order!numeric ($<$, $<=$, $=$, $>=$, $>$)}\index{operator!numeric order ($<$, $<=$, $=$, $>=$, $>$)}\index{number!order ($<$, $<=$, $=$, $>=$, $>$)}\seeonlyindex{greater or equal ($>=$)}{predicate}greater
than or equal to (\textsf{>=}) both $x$ and $y$.
Use Proof Pad to run tests of the property.}

\exer {Define a DoubleCheck property to test the distributive law
of arithmetic (\{distributive law\}, figure~\ref{fig-02-01}, page \pageref{fig-02-01}).
Use Proof Pad to run your test.}

\exer {Define DoubleCheck properties to test consistency between clock
arithmetic and ordinary arithmetic as described in
box~\ref{modular-arithmetic} (page \pageref{modular-arithmetic}).
Use Proof Pad to run your tests.}

\exer {The ACL2 operator
\index{operator, by name!reverse (elements in list)}\seeonlyindex{reverse}{operator}\textsf{reverse}
delivers a list whose elements are in the reverse order of those in its operand.
For example \textsf{(reverse (list 1 2 3))} is \textsf{(list 3 2 1)}.
Find an equation that expresses the value of \textsf{(reverse (append $xs$ $ys$))}
in terms of \textsf{(reverse $xs$)} and \textsf{(reverse $ys$)}.
Define a property based on your equation and
use Proof Pad to test it.}

\end{exercises}

\begin{aside}{floor-ceiling-ops-brackets}{Floor and Ceiling Operators, Floor and Ceiling Brackets}
ACL2 provides two operators,
\textsf{ceiling} and \textsf{floor}, that divide one number by another,
then round the result up or down to an integer value.
In algebraic formulas, special brackets known as
floor and ceiling brackets indicate the direction of rounding.
Ceiling brackets (like square brackets with the bottom chopped off)
indicate the next integer that is the same as or larger than
the enclosed value.
Floor brackets (square brackets without the top part)
go in the other direction, down to the next integer
that is the same as or smaller than the enclosed value.
\vspace{2mm}
\begin{center}
\index{operator, by name!ceiling (divide, round up)}\seeonlyindex{ceiling}{operator}\index{ceiling brackets}\index{brackets!ceiling}\index{operator, by name!floor (divide, round down)}\seeonlyindex{floor}{operator}\index{division!floor, mod}\index{floor brackets}\index{brackets!floor}
\begin{tabular}{lll}
\textsf{(floor $x$ $y$)}  & $= \lfloor x \div y \rfloor$ \emph{quotient rounded down} \vspace{1mm}\\
\textsf{(ceiling $x$ $y$)}& $= \lceil x  \div y \rceil $ \emph{quotient rounded up}   \\
\end{tabular}
\end{center}
%\caption{Floor and Ceiling Operators, Floor and Ceiling Brackets}
%\label{floor-def}
%\label{ceiling-def}
%\label{floor-ceiling-ops-brackets}
\end{aside}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "book"
%%% End:
